New cardinality estimation algorithms for HyperLogLog sketches

### 2017-02-03 paper/paper.tex
minor changes

\section{Introduction}
Counting the number of distinct elements in a data stream or large datasets is a common problem in big data processing. In principle, finding the number of distinct elements $\symCardinality$ with a maximum relative error $\symError$  in a data stream requires $\Omega(\symCardinality)$ space \cite{Alon1999}. However, probabilistic algorithms that achieve the requested accuracy only with high probability are able to drastically reduce space requirements. Many different probabilistic algorithms have been developed over the past two decades \cite{Metwally2008,Ting2014} until a theoretically optimal algorithm was finally found \cite{Kane2010}. Although this algorithm achieves the optimal space complexity of $\Omega(\symError^{-2}+\log \symCardinality)$ \cite{Alon1999, Indyk2003}, it is not very efficient in practice \cite{Ting2014}.

More practicable and already widely used for many applications is the HyperLogLog algorithm \cite{Flajolet2007} with a near-optimal space complexity $\Omega(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$. It has the nice property that partial results can be easily merged, which is a necessity for distributed environments. Unfortunately, the originally proposed estimation method has some problems to guarantee the same accuracy over the full cardinality range. Therefore, a couple of variants have been developed to correct the original estimate by empirical means \cite{Heule2013,Rhodes2015,Sanfilippo2014}.

@@

\subsection{Outline}
This paper presents new algorithms for cardinality estimation from HyperLogLog sketches. In \cref{sec:hyperloglog_data_structure} we start with an introduction to HyperLogLog sketches and the corresponding update algorithm. We also discuss how the resulting state can be statistically described and approximated by a Poisson model.
In \cref{sec:cardinality_estimation} we describe the original cardinality estimation algorithm, its shortcomings at low and large cardinalities, and previous approaches to overcome them. We present a simple derivation of the original raw estimator and analyze the root cause for its limited operating range. We derive an improved version of the raw estimator, which leads to a new fast cardinality estimation algorithm that works over the full cardinality range as demonstrated by extensive simulations.
In \cref{sec:max_likelihood_estimation} we derive a second even more precise cardinality estimation algorithm based on the maximum likelihood principle which is again verified by simulations. As presented in \cref{sec:cardinality_estimation_set_intersections} the same approach can be generalized to two HyperLogLog sketches, which allows result cardinality estimation of set operations like intersections or complements. The simulation results show that the estimates are significantly better than those obtained using the inclusion-exclusion principle.
Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential application as locality-sensitve hashing algorithm before we conclude in \cref{sec:conclusion}.

\section{HyperLogLog data structure}
\label{sec:hyperloglog_data_structure}
The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers. For performance reasons the number of registers $\symNumReg$ is chosen to be a power of 2, $\symNumReg = 2^\symPrecision$. $\symPrecision$ is the precision parameter that directly controls the relative error which scales like $1/\sqrt{\symNumReg}$ \cite{Flajolet2007}. All registers start with zero initial value. Each element insertion potentially increases the value of one of these registers. The maximum value a register can reach is a natural bound given either by the output size of the used hash algorithm or the space that is reserved for a single register. Common implementations allocate up to 8 bits per register.

\subsection{Data element insertion}
\label{sec:data_element_insertion}
In order to insert a data element into a HyperLogLog data structure a hash value is calculated. The leading $\symPrecision$ bits of the hash value are used to select one of the $2^\symPrecision$ registers. Among the next following $\symRegRange$ bits, the position of the first 1-bit is determined which is a value in the range $[1,\symRegRange+1]$. The value $\symRegRange+1$ is used, if all $\symRegRange$ bits are zeros. If the position of the first 1-bit exceeds the current value of the selected register, the register value is replaced. \cref{alg:insert} shows the update procedure for inserting a data element into the HyperLogLog sketch.

@@

\subsection{Poisson approximation}
\label{sec:poisson_approximation}
Due to the statistical dependence of the register values, the probability mass function \eqref{equ:multinomialProbabilityMass} makes further analysis difficult. Therefore, a Poisson model can be used \cite{Flajolet2007}, which assumes that the cardinality itself is distributed according to a Poisson distribution
\begin{equation}
\symCardinality \sim \text{Poisson}(\symPoissonRate).
\end{equation}
Under the Poisson model the distribution of the register values is
\begin{align}
\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symPoissonRate) 
&= 
\sum_{\symCardinality=0}^\infty \symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\label{equ:poissonization}
\\
&= 
\sum_{\symCardinality_1=0}^\infty
\cdots
\sum_{\symCardinality_\symNumReg=0}^\infty
\prod_{\symIndexI=1}^\symNumReg
\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexI}}{\symCardinality_\symIndexI!\symNumReg^{\symCardinality_\symIndexI}}
\nonumber\\
&= 
\prod_{\symIndexI=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\nonumber\\
&= 
\prod_{\symRegVal=0}^{\symRegRange+1} \left(
\sum_{\symCardinality=0}^\infty
\symRegProbability(\symRegVal\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\right)^{\!\!\symCount_\symRegVal}
\nonumber\\
\label{equ:poisson_pmf}
&=
e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\!\symCount_\symRegVal}\right)
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\!\symCount_{\symRegRange+1}}.
\end{align}
Here $\symCount_\symRegVal$ denotes the multiplicity of value $\symRegVal$  in the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$.
The final factorization shows that under the Poisson model the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed. The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is given by
\begin{equation}
\label{equ:register_value_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
=
\begin{cases}
0 & \symRegVal < 0 \\
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
1 & \symRegVal > \symRegRange.
\end{cases}
\end{equation}

\subsection{Depoissonization}
\label{sec:depoissonization}
Due to the simpler probability mass function, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\boldsymbol{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ than for the cardinality $\symCardinality$ under the fixed-size model \eqref{equ:multinomialProbabilityMass}. Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model. Assume we have found an unbiased estimator for the Poisson rate
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = \symPoissonRate
\quad
\text{for all $\symPoissonRate\geq 0$}.
\end{equation}
We know from \eqref{equ:poissonization} 
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = 
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\end{equation}
and therefore
\begin{equation}
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
=
\symPoissonRate
\end{equation}
holds for all $\symPoissonRate\geq0$. The unique solution of this equation is given by
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symCardinality) = \symCardinality.
\end{equation}
Hence, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$, which motivates us to use $\symPoissonRateEstimate$ directly as estimator for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$. As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.

\section{Original cardinality estimation method}
\label{sec:cardinality_estimation}
The original cardinality estimator \cite{Flajolet2007} is based on the idea that the number of distinct element insertions a register needs to reach the value $\symRegVal$ is proportional to $\symNumReg  2^{\symRegVal}$. Given that, a rough cardinality estimate can be obtained by averaging the values $\lbrace \symNumReg 2^{\symRegValVariate_1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg}\rbrace$. 
In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator which is given by
\begin{equation}
\label{equ:raw_estimator}
\symCardinalityRawEstimate
=
\alpha_\symNumReg
\frac{\symNumReg}
{\frac{1}{\symNumReg 2^{\symRegValVariate_1}}+\ldots+\frac{1}{\symNumReg2^{\symRegValVariate_\symNumReg}}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexI=1}^{\symNumReg}2^{-\symRegValVariate_\symIndexI}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal 2^{-\symRegVal}}.
\end{equation}
Here $\alpha_\symNumReg$ is a bias correction factor which was derived for a given number of registers $\symNumReg$ as \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\!\left(
\frac{2+\symX}{1+\symX}
\right)
\right)^{\!\symNumReg}
d\symX
\right)^{\!\!-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} 
= 0.72134752\ldots,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:raw_estimator}, because the additional bias is negligible compared to the overall estimation error.

@@

If performance matters $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$
can also be precalculated for all possible values of $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$. As their value range is $\{0,1,\ldots, \symNumReg\}$, the function values can be kept in lookup tables of size $\symNumReg+1$. In this way a complete branch-free cardinality estimation can be realized. It is also thinkable to calculate $\symLargeCorrectionFunc$ using the approximation $\symLargeCorrectionFunc(\symX)\approx -\symAlpha_\infty/\log(\symX)-\symSmallCorrectionFunc(\symX)$ which can be obtained from \eqref{equ:sigma_tau_relationship}. The advantage is that the calculation of $\symSmallCorrectionFunc$ and the additional logarithm is slightly faster than the calculation of $\symSmallCorrectionFunc(\symX)$. 
However, for arguments close to 1, where $1/\log(\symX)$ and $\symSmallCorrectionFunc(\symX)$ are both very large while their difference is very small, special care is needed to avoid numerical cancellation.

@@

\subsection{Experimental setup}
\label{sec:experimental_setup}
In order to verify the new estimation algorithm, we generated \num{10000} different HyperLogLog sketches and filled each of them with 50 billion unique elements. During element insertion, we took snapshots of the sketch state by storing the multiplicity vector at predefined cardinality values, which were roughly chosen according to a geometric series with ratio \num{1.01}. In order to achieve that in reasonable time for different HyperLogLog parameter combinations  $(\symPrecision,\symRegRange)$ we have been interested in, we applied a couple of optimizations.

First, we assumed a uniform hash function. This allows us to simply generate a random number instead of getting a unique element and calculating its hash value. We used the Mersenne Twister random number generator with a state size of \num{19937} bits from the C++ standard library. 
Second, we used \cref{alg:insert_fast} for insertions, which does basically the same as \cref{alg:insert}, but additionally keeps track of the multiplicity vector and the minimum register value. In this way the expensive register scan needed to obtain the multiplicity vector as described by \cref{alg:sufficient_statistic} can be avoided. Furthermore, the knowledge of the minimum register value can be used to abort the insertion procedure early, if the hash value is not able to increment any register value. Especially at large cardinalities this saves a great number of register accesses and also reduces the number of cache misses. For example, if $\symRegValVariate_\text{min}=1$ only half of all insertions will access a register. If $\symRegValVariate_\text{min}=2$ it is already just a quarter, and so forth. 
Finally, instead of repeating the simulation for different HyperLogLog parameters, we did that only once using the quite accurate setting $\symPrecision=22$ and $\symRegRange=42$.  Whenever taking a snapshot, we reduced the sketch state exploiting \cref{alg:compress} for desired parameters before calculating the multiplicity vector using \cref{alg:sufficient_statistic}.

@@

The new estimation algorithm also works well for other HyperLogLog configurations. First we considered configurations using a 32-bit hash function ($\symPrecision + \symRegRange = 32$). The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=22$ are shown in \cref{fig:raw_corrected_estimation_error_8_24}, \cref{fig:raw_corrected_estimation_error_16_16}, and \cref{fig:raw_corrected_estimation_error_22_10}, respectively. As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision. Again, the new algorithm gives essentially unbiased estimates. Only for very high precisions, an oscillating bias becomes apparent (compare \cref{fig:raw_corrected_estimation_error_22_10}), that is caused by approximating the periodic function $\symPowerSeriesFunc$ by a constant (see \cref{sec:derivation_raw_estimator}).

@@

The results show that the execution times are nearly constant for $\symRegRange=52$. Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed due to the small probability of saturated registers, $\symCountVariate_{53}=0$ usually holds for realistic cardinalities.

For the case $\symRegRange=20$ with precalculated correction values the computation time is again -- as expected -- independent of the cardinality. The faster computation times compared to the $\symRegRange=52$ case can be explained by the much smaller dimension of the multiplicity vector which is equal to $\symRegRange+2$. If the functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are calculated on demand, the execution times for larger cardinalities are doubled. This comes from the fact that the calculation of $\symLargeCorrectionFunc$ is much more expensive than that of $\symSmallCorrectionFunc$, because it requires more iterations and involves square root evaluations. If $2^{\symPrecision+\symRegRange}$ is greater than the maximum expected cardinality value, the number of saturated registers is negligible ($\symCountVariate_{\symRegRange+1}\ll \symNumReg$) and the calculation of $\symLargeCorrectionFunc$ can be omitted at all, because $\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg) \approx 0$ vanishes in this case.

>>>>

\section{Introduction}
Counting the number of distinct elements in a data stream or large datasets is a common problem in big data processing. In principle, finding the number of distinct elements $\symCardinality$ with a maximum relative error $\symError$  in a data stream requires $\symBigO(\symCardinality)$ space \cite{Alon1999}. However, probabilistic algorithms that achieve the requested accuracy only with high probability are able to drastically reduce space requirements. Many different probabilistic algorithms have been developed over the past two decades \cite{Metwally2008,Ting2014} until a theoretically optimal algorithm was finally found \cite{Kane2010}. Although this algorithm achieves the optimal space complexity of $\symBigO(\symError^{-2}+\log \symCardinality)$ \cite{Alon1999, Indyk2003}, it is not very efficient in practice \cite{Ting2014}.

More practicable and already widely used for many applications is the HyperLogLog algorithm \cite{Flajolet2007} with a near-optimal space complexity $\symBigO(\symError^{-2} \log\log\symCardinality +\log \symCardinality)$. It has the nice property that partial results can be easily merged, which is a necessity for distributed environments. Unfortunately, the originally proposed estimation method has some problems to guarantee the same accuracy over the full cardinality range. Therefore, a couple of variants have been developed to correct the original estimate by empirical means \cite{Heule2013,Rhodes2015,Sanfilippo2014}.

@@

\subsection{Outline}
This paper presents new algorithms to extract cardinality information from HyperLogLog sketches. In \cref{sec:hyperloglog_data_structure} we start with an introduction to HyperLogLog sketches and the corresponding update algorithm. We also discuss how the resulting state can be statistically described and approximated by a Poisson model.
In \cref{sec:cardinality_estimation} we describe the original cardinality estimation algorithm, its shortcomings at low and large cardinalities, and previous approaches to overcome them. We present a simple derivation of the original raw estimator and analyze the root cause for its limited operating range. We derive an improved version of the raw estimator, which leads to a new fast cardinality estimation algorithm that works over the full cardinality range as demonstrated by extensive simulations.
In \cref{sec:max_likelihood_estimation} we derive a second even more precise cardinality estimation algorithm based on the maximum likelihood principle which is again verified by simulations. As presented in \cref{sec:cardinality_estimation_set_intersections} the same approach can be generalized to two HyperLogLog sketches, which allows result cardinality estimation of set operations like intersections or complements. The simulation results show that the estimates are significantly better than those of the conventional approach using the inclusion-exclusion principle.
Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential application as locality-sensitve hashing algorithm before we conclude in \cref{sec:conclusion}.

\section{HyperLogLog data structure}
\label{sec:hyperloglog_data_structure}
The HyperLogLog algorithm collects information of incoming elements into a 
very compact sketching data structure, that finally allows the estimation of the number of distinct elements. The data structure consists of $\symNumReg = 2^\symPrecision$ registers whose number is chosen to be a power of two for performance reasons. $\symPrecision$ is the precision parameter that directly controls the relative estimation error which scales like $1/\sqrt{\symNumReg}$ \cite{Flajolet2007}. All registers start with zero initial value. Each element insertion potentially increases the value of one of these registers. The maximum value a register can reach is a natural bound given either by the output size of the used hash algorithm or the space that is reserved for a single register. Common implementations allocate up to 8 bits per register.

\subsection{Data element insertion}
\label{sec:data_element_insertion}
The insertion of a data element into a HyperLogLog data structure requires the calculation of a $(\symPrecision+\symRegRange)$-bit hash value. The leading $\symPrecision$ bits of the hash value are used to select one of the $2^\symPrecision$ registers. Among the next following $\symRegRange$ bits, the position of the first 1-bit is determined which is a value in the range $[1,\symRegRange+1]$. The value $\symRegRange+1$ is used, if all $\symRegRange$ bits are zeros. If the position of the first 1-bit exceeds the current value of the selected register, the register value is replaced. \cref{alg:insert} shows the update procedure for inserting a data element into the HyperLogLog sketch.

@@

\subsection{Poisson approximation}
\label{sec:poisson_approximation}
Due to the statistical dependence of the register values, the probability mass function \eqref{equ:multinomialProbabilityMass} makes further analysis difficult. Therefore, a Poisson model can be used \cite{Flajolet2007}, which assumes that the cardinality itself is distributed according to a Poisson distribution
\begin{equation}
\symCardinality \sim \text{Poisson}(\symPoissonRate).
\end{equation}
Under the Poisson model the distribution of the register values is
\begin{align}
\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symPoissonRate) 
&= 
\sum_{\symCardinality=0}^\infty \symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\label{equ:poissonization}
\\
&= 
\sum_{\symCardinality_1=0}^\infty
\cdots
\sum_{\symCardinality_\symNumReg=0}^\infty
\prod_{\symIndexI=1}^\symNumReg
\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexI}}{\symCardinality_\symIndexI!\symNumReg^{\symCardinality_\symIndexI}}
\nonumber\\
&= 
\prod_{\symIndexI=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\nonumber\\
&= 
\prod_{\symRegVal=0}^{\symRegRange+1} \left(
\sum_{\symCardinality=0}^\infty
\symRegProbability(\symRegVal\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\right)^{\!\!\symCount_\symRegVal}
\nonumber\\
\label{equ:poisson_pmf}
&=
e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\!\symCount_\symRegVal}\right)
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\!\symCount_{\symRegRange+1}}.
\end{align}
Here $\symCount_\symRegVal$ denotes the multiplicity of value $\symRegVal$  in the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$.
The final factorization shows that the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed under the Poisson model. The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is given by
\begin{equation}
\label{equ:register_value_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
=
\begin{cases}
0 & \symRegVal < 0 \\
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
1 & \symRegVal > \symRegRange.
\end{cases}
\end{equation}

\subsection{Depoissonization}
\label{sec:depoissonization}
Due to the simpler probability mass function \eqref{equ:poisson_pmf}, it is easier to find an estimator $\symPoissonRateEstimate = \symPoissonRateEstimate(\boldsymbol{\symRegValVariate})$ for the Poisson rate $\symPoissonRate$ than for the cardinality $\symCardinality$ under the fixed-size model \eqref{equ:multinomialProbabilityMass}. Depoissonization \cite{Jacquet1998} finally allows to translate the estimates back to the fixed-size model. Assume we have found an unbiased estimator for the Poisson rate
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = \symPoissonRate
\quad
\text{for all $\symPoissonRate\geq 0$}.
\end{equation}
We know from \eqref{equ:poissonization} 
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symPoissonRate) = 
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\end{equation}
and therefore
\begin{equation}
\sum_{\symCardinality=0}^\infty \symExpectation(\symPoissonRateEstimate\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
=
\symPoissonRate
\end{equation}
holds for all $\symPoissonRate\geq0$. The unique solution of this equation is given by
\begin{equation}
\symExpectation(\symPoissonRateEstimate\vert\symCardinality) = \symCardinality.
\end{equation}
Hence, the unbiased estimator $\symPoissonRateEstimate$ conditioned on $\symCardinality$ is also an unbiased estimator for $\symCardinality$, which motivates us to use $\symPoissonRateEstimate$ directly as estimator for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$. As simulation results will show later, the Poisson approximation works well over the entire cardinality range, even for estimators that are not exactly unbiased.

\section{Original cardinality estimation method}
\label{sec:cardinality_estimation}
The original cardinality estimator \cite{Flajolet2007} is based on the idea that the number of distinct element insertions a register needs to reach the value $\symRegVal$ is proportional to $\symNumReg  2^{\symRegVal}$. Given that, a rough cardinality estimate can be obtained by averaging the values $\lbrace \symNumReg 2^{\symRegValVariate_1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg}\rbrace$. 
In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator which is given by
\begin{equation}
\label{equ:raw_estimator}
\symCardinalityRawEstimate
=
\alpha_\symNumReg
\frac{\symNumReg}
{\frac{1}{\symNumReg 2^{\symRegValVariate_1}}+\ldots+\frac{1}{\symNumReg2^{\symRegValVariate_\symNumReg}}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexI=1}^{\symNumReg}2^{-\symRegValVariate_\symIndexI}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal 2^{-\symRegVal}}.
\end{equation}
Here $\alpha_\symNumReg$ is a bias correction factor which was derived for a given number of registers $\symNumReg$ as \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\!\left(
\frac{2+\symX}{1+\symX}
\right)
\right)^{\!\symNumReg}
d\symX
\right)^{\!\!-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} 
= 0.72134752\ldots
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:raw_estimator}, because the additional bias is negligible compared to the overall estimation error.

@@

If performance matters $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$
can also be precalculated for all possible values of $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$. As their value range is $\{0,1,\ldots, \symNumReg\}$, the function values can be kept in lookup tables of size $\symNumReg+1$. In this way a complete branch-free cardinality estimation can be realized. It is also thinkable to calculate $\symLargeCorrectionFunc$ using the approximation $\symLargeCorrectionFunc(\symX)\approx \symAlpha_\infty/\log(1/\symX)-\symSmallCorrectionFunc(\symX)$ which can be obtained from \eqref{equ:sigma_tau_relationship}. The advantage is that the calculation of $\symSmallCorrectionFunc$ and the additional logarithm is slightly faster than the calculation of $\symLargeCorrectionFunc$. 
However, for arguments close to 1, where $\symAlpha_\infty/\log(1/\symX)$ and $\symSmallCorrectionFunc(\symX)$ are both very large while their difference is very small, special care is needed to avoid numerical cancellation.

@@

\subsection{Experimental setup}
\label{sec:experimental_setup}
To verify the new estimation algorithm, we generated \num{10000} different HyperLogLog sketches and filled each of them with 50 billion unique elements. During element insertion, we took snapshots of the sketch state by storing the multiplicity vector at predefined cardinality values, which were roughly chosen according to a geometric series with ratio \num{1.01}. In order to achieve that in reasonable time for all the different HyperLogLog parameter combinations  $(\symPrecision,\symRegRange)$ we have been interested in, we applied a couple of optimizations.

First, we assumed a uniform hash function. This allows us to simply generate random numbers instead of creating unique elements and calculating their hash values. We used the Mersenne Twister random number generator with a state size of \num{19937} bits from the C++ standard library. 
Second, we used \cref{alg:insert_fast} for insertions, which does basically the same as \cref{alg:insert}, but additionally keeps track of the multiplicity vector and the minimum register value. In this way the expensive register scan needed to obtain the multiplicity vector as described by \cref{alg:sufficient_statistic} can be avoided. Furthermore, the knowledge of the minimum register value can be used to abort the insertion procedure early, if the hash value is not able to increment any register value. Especially at large cardinalities this saves a great number of register accesses and also reduces the number of cache misses. For example, if $\symRegValVariate_\text{min}=1$ only half of all insertions will access a register. If $\symRegValVariate_\text{min}=2$ it is already just a quarter, and so forth. 
Finally, instead of repeating the simulation for different HyperLogLog parameters, we did that only once using the quite accurate setting $\symPrecision=22$ and $\symRegRange=42$.  Whenever taking a snapshot, we reduced the sketch state exploiting \cref{alg:compress} for desired parameters before calculating the multiplicity vector using \cref{alg:sufficient_statistic}.

@@

The new estimation algorithm also works well for other HyperLogLog configurations. First we considered configurations using a 32-bit hash function ($\symPrecision + \symRegRange = 32$). The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=22$ are shown in \cref{fig:raw_corrected_estimation_error_8_24}, \cref{fig:raw_corrected_estimation_error_16_16}, and \cref{fig:raw_corrected_estimation_error_22_10}, respectively. As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision parameter $\symPrecision$. Again, the new algorithm gives essentially unbiased estimates. Only for very high precisions, an oscillating bias becomes apparent (compare \cref{fig:raw_corrected_estimation_error_22_10}), that is caused by approximating the periodic function $\symPowerSeriesFunc$ by a constant (see \cref{sec:derivation_raw_estimator}).

@@

The results show that the execution times are nearly constant for $\symRegRange=52$. Using a lookup table makes not much difference, because the on-demand calculation for  $\symSmallCorrectionFunc$ is very fast and the calculation of $\symLargeCorrectionFunc$ is rarely needed due to the small probability of saturated registers, $\symCountVariate_{53}=0$ usually holds for realistic cardinalities. If lookup tables are used, the computation time is as expected independent of the cardinality also for the case $\symRegRange=20$. The faster computation times for $\symRegRange=20$ compared to the $\symRegRange=52$ case can be explained by the much smaller dimension of the multiplicity vector which is equal to $\symRegRange+2$. In contrast, if functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are calculated on demand, executions take significantly more time for larger cardinalities. The reason is that beginning at cardinality values in the order of \num{100000} the probability of saturated registers increases. This makes the calculation of $\symLargeCorrectionFunc$ necessary, which is much more expensive than that of $\symSmallCorrectionFunc$, because it requires more iterations and involves square root evaluations. However, the calculation of 
$\symLargeCorrectionFunc$ could be avoided at all, if the HyperLogLog parameters are appropriately chosen. If $2^{\symPrecision+\symRegRange}$ is much larger than the maximum expected cardinality value, the number of saturated registers will be negligible, $\symCountVariate_{\symRegRange+1}\ll \symNumReg$. The calculation of $\symLargeCorrectionFunc$ could be omitted in this case, because $\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg) \approx 0$.
